Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-28 manuscript/manuscript.tex
Move the description of grids and shells to the top
The description of each grid and shell model was in the linear density
section but we them for the other functions as well. I moved them up to
the beginning of the section and added some more justification about
why we chose these values (as requested by the reviewers).
I also simplified the formatting of the shell models table. I think this
is simpler and the caption can be a bit shorter because of it.

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
We derive an expression for such gravitational potential in case the spherical shell has
an arbitrary density function in depth (Eq.~\ref{eq:shell-pot}), and obtained the
expressions in case of linear and exponential density (see Appendix~\ref{sec:shell}).

We perform comparisons between the analytical solutions for the spherical shell and the
numerical integration results for linear, exponential and trigonometric density
functions.
From these results, we generalize default values for $D$ and $\delta$ that ensure a
numerical error lower than 0.1\% of the spherical shell values.

\begin{table}
\caption{
    Description of the tesseroid models used to build spherical shells and characterize
    the accuracy of the numerical integration.
    The first column indicates the thickness of the shell model and the next three
    describe the shape of the tesseroid model: each one specifies the number of
    tesseroids on the corresponding dimension (radial, latitudinal, longitudinal).
    The last two columns specify the latitudinal and longitudinal size of each
    tesseroid that compose the spherical shell model.
    The outer radius of every shell model is equal to the mean Earth radius (6378.137
    km), while the inner radius is determined by its thickness.
}
\label{tab:shell-models}
\begin{tabular}{cccccc}
    Thickness & Radial & Latitudinal & Longitudinal & $L_\phi$ & $L_\lambda$ \\ \hline
    100m   & 1 & 6 & 12 & 30$^\circ$ & 30$^\circ$ \\
    1km    & 1 & 6 & 12 & 30$^\circ$ & 30$^\circ$ \\
    10km   & 1 & 6 & 12 & 30$^\circ$ & 30$^\circ$ \\
    100km  & 1 & 6 & 12 & 30$^\circ$ & 30$^\circ$ \\
    1000km & 1 & 6 & 12 & 30$^\circ$ & 30$^\circ$ \\
\end{tabular}
\end{table}

\begin{table}
\caption{
    Description of the computation grids used to characterize the accuracy of the
    numerical integration.
    Grid height is defined above the mean Earth radius.
}
\label{tab:grids}
\begin{tabular}{lccc}
    Name & Grid size & Grid region (degrees) & Grid height (km)
    \\ \hline
    Pole      & $0.1^\circ \times 0.1^\circ$ &   0E /   1E / 89N / 90N & 0   \\
    Equator   & $0.1^\circ \times 0.1^\circ$ &   0E /   1E /  0N / 1N  & 0   \\
    Global    & $ 10^\circ \times  10^\circ$ & 180W / 180E / 90S / 90N & 0   \\
    Satellite & $ 10^\circ \times  10^\circ$ & 180W / 180E / 90S / 90N & 260 \\
\end{tabular}
\end{table}

>>>>

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
We derive an expression for the gravitational potential of a spherical shell with
an arbitrary density function in depth (Eq.~\ref{eq:shell-pot}),
from which we obtained
expressions for linear, exponential, and sinusoidal density functions
(see Appendix~\ref{sec:shell}).

In order to compare the numerical results with the analytical solution we must build
spherical shell models out of tesseroids.
We divide the spherical shell along the latitudinal and longitudinal directions to
obtain a shell model made out of $6 \times 12 = 72$ tesseroids of size
$30^\circ \times 30^\circ$.
In order to asses the density-based discretization in the radial dimension,
we use spherical shell models with different thicknesses (Table~\ref{tab:shell-models}).
The thickness values were chosen to represent a range of applications, from topographic
to lithospheric scale models.
Because the amount of tesseroid divisions in the adaptive discretization will be
proportional to the tesseroid size (Eq.~\ref{eq:condition}),
some of these configurations represent worse-case scenarios.
Most practical applications will use tesseroids smaller than
$30^\circ \times 30^\circ \times 1000\ \text{km}$.

The differences between the analytical and numerical solutions could be calculated on a
single computation point due to the rotational symmetry of the spherical shell.
However, the numerical results depend on the relative location between the computation
point and the tesseroid \citep{Ku1977, Asgharzadeh2007, Uieda2016}.
We account for this effect by calculating the differences on regular grids and keeping
only the maximum absolute difference.
All calculations are repeated for four different computation grids
(Table~\ref{tab:grids}):
a local grid at the pole, a local grid at the equator, a global grid at zero height
above the shell, and a global grid at a height of 260 km (representing the nominal
height of the GOCE satellite).
These grids span a broad scenario of use cases and ensure an acceptable accuracy on each
one of them.

We perform comparisons between the analytical solutions for the spherical shell and the
numerical integration results for linear, exponential, and sinusoidal density
functions.
The comparisons are repeated for all combinations of tesseroid models in
Table~\ref{tab:shell-models} and computation grids in Table~\ref{tab:grids}.
From these results, we generalize optimal values for $D$ and $\delta$ that ensure a
numerical error lower than 0.1\% of the spherical shell values for most use cases.

\begin{table}
\caption{
    Description of the tesseroid models used to build spherical shells and characterize
    the accuracy of the numerical integration.
    The outer radius ($R_2$) of every shell model is equal to the mean Earth radius
    (6378.137 km), while the inner radius ($R_1$) is determined by its thickness.
    The horizontal dimensions of the tesseroids and the total number of
    tesseroids in the shell model are given in the latitudinal and longitudinal
    dimensions, respectively.
}
\label{tab:shell-models}
\begin{tabular}{rccccc}
    Thickness & Tesseroid size & Number of tesseroids \\ \hline
    0.1 km  & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    1 km    & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    10 km   & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    100 km  & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
    1000 km & $30^\circ \times 30^\circ$ & $6 \times 12 = 72$ \\
\end{tabular}
\end{table}

\begin{table}
\caption{
    Description of the computation grids used to characterize the accuracy of the
    numerical integration.
    Grid height is defined above the mean Earth radius.
}
\label{tab:grids}
\begin{tabular}{lccc}
    Name & Grid spacing & Grid region (degrees) & Grid height (km)
    \\ \hline
    Pole      & $0.1^\circ$ &   0E /   1E / 89N / 90N & 0   \\
    Equator   & $0.1^\circ$ &   0E /   1E /  0N / 1N  & 0   \\
    Global    & $ 10^\circ$ & 180W / 180E / 90S / 90N & 0   \\
    Satellite & $ 10^\circ$ & 180W / 180E / 90S / 90N & 260 \\
\end{tabular}
\end{table}
